---
title: "Adding a regression line on a ggplot2"
output: 
  distill::distill_article:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

This short post will show how to add a regression line to a [ggplot2](https://ggplot2.tidyverse.org/) chart. We start by loading in the `USpop` data found in the [second lab](labs-02.html).

```{r, echo=FALSE}
USpop <- read.csv(here::here("data", "USpop.csv"))
```

```{r, eval=FALSE}
USpop <- read.csv("USpop.csv")
```

Next, we create a simple linear regression model

```{r}
my_fit <- lm(Population ~ Year, data = USpop)

my_fit
```

Let us start with a simple scatter chart using ggplot2

```{r}
library(ggplot2)

ggplot(USpop, aes(Year, Population)) +
  geom_point()
```

Adding a regression line to this chart can be done in a couple of different ways.

# Using `geom_smooth()`

The first way is by far the most simple. Using `geom_smooth()` fits a line directly to the data inside ggplot itself, if we specify `method = "lm` and `formula = y ~ x` then we force that line to a simple linear regression. Setting `se = FALSE` hides the confidence intervals.

```{r}
ggplot(USpop, aes(Year, Population)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
```

This method while pleasing doesn't answer the question at hand. It doesn't let us add our already fitted model, but instead fits a model itself.

# using `broom::augment()`

You can use the [broom]() package to extract various information from the model fit.

```{r}
library(broom)
augment(my_fit)
```

Using this data along with `geom_line()` allows us to all a fitted line on top of our ggplot

```{r}
ggplot(USpop, aes(Year, Population)) +
  geom_point() +
  geom_line(data = augment(my_fit), 
            aes(x = Year, y = .fitted))
```

The main downside to this way is that it does not easily extrapolate outside the range so we zoom out the line doesn't expand

```{r}
ggplot(USpop, aes(Year, Population)) +
  geom_point() +
  geom_line(data = augment(my_fit), 
            aes(x = Year, y = .fitted)) +
  lims(x = c(1700, 2100), y = c(-100, 500))
```
# `coef()` and `geom_abline()`

Lastly, we can extract the parameters estimates directly with `coef()` and use these to add a single line with `geom_abline()`

```{r}
my_coef <- coef(my_fit)
my_coef
```

```{r}
ggplot(USpop, aes(Year, Population)) +
  geom_point() +
  geom_abline(intercept = my_coef[1], slope = my_coef[2])
```

Since `geom_abline()` add a line you can zoom out and still see the line

```{r}
ggplot(USpop, aes(Year, Population)) +
  geom_point() +
  geom_abline(intercept = my_coef[1], slope = my_coef[2]) +
  lims(x = c(1700, 2100), y = c(-100, 500))
```
