---
title: "Lab - week 5"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

Based on **Applied Statistics with R** by David Dalpiaz (https://github.com/daviddalpiaz/appliedstats)

# `cars` Example

We will use the small dataset `cars` which comes with R and it availiable right away

```{r}
cars
```

We want too examine the relationship between the speed of a car (`speed`) and its stopping distance (`dist`).

a. Start by creating a scatter plot of `dist` and `speed`. Describe what you see.

```{r, eval=FALSE}
plot(dist ~ speed, data = cars)
```

b. Fit a simple linear regression model on `cars` with `dist` as the response variable and `speed` as the predictor variable. Plot it on top of the scatter plot. How does it fit?

```{r, eval=FALSE}
stop_dist_model <- lm(dist ~ speed, data = cars)
plot(dist ~ speed, data = cars)
abline(stop_dist_model)
```

c. Examine the model fit with `summary()`. Look at the different results from `summary()` using `names()`.

d. Verify the value of the $t$ test statistic for $\b_1$ and the two-sided p-value associated with that test statistic.

e. Use the `confint()` function to obain the confidence interval for $\beta_0$ and $\beta_1$.

f. Verify the calculations that R is performing for the $\beta_1$ interval.
