---
title: "Extensions of the Linear Model"
subtitle: "AU STAT-615"
author: "Emil Hvitfeldt"
date: "2021-04-07"
output:
  xaringan::moon_reader:
    css: ["theme.css", "default"]
    lib_dir: libs
    nature:
      beforeInit: "macros.js"
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: [center, middle]
---

```{r include=FALSE}
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
lines <- options$output.lines
if (is.null(lines)) {
  return(hook_output(x, options))  # pass to default hook
}
x <- unlist(strsplit(x, "\n"))
more <- "..."
if (length(lines) == 1) {        # first n lines
  if (length(x) > lines) {
    # truncate the output, but add ....
    x <- c(head(x, lines), more)
  }
} else {
  x <- c(more, x[lines], more)
}
# paste these lines together
x <- paste(c(x, ""), collapse = "\n")
hook_output(x, options)
})
knit_hooks$set(output = function(x, options) {
# this hook is used only when the linewidth option is not NULL
if (!is.null(n <- options$linewidth)) {
  x = knitr:::split_lines(x)
  # any lines wider than n should be wrapped
  if (any(nchar(x) > n)) x = strwrap(x, width = n)
  x = paste(x, collapse = '\n')
}
hook_output(x, options)
})

opts_chunk$set(
echo = TRUE,
fig.width = 7, 
fig.align = 'center',
fig.asp = 0.618, # 1 / phi
out.width = "700px"
)
```

```{r, echo = FALSE, message=FALSE}
library(sass)
sass(sass_file("theme.sass"), output = "theme.css")
library(tibble)
library(dplyr)
library(ggplot2)
```

$$\require{color}\definecolor{orange}{rgb}{1, 0.603921568627451, 0.301960784313725}$$
$$\require{color}\definecolor{blue}{rgb}{0.301960784313725, 0.580392156862745, 1}$$
$$\require{color}\definecolor{pink}{rgb}{0.976470588235294, 0.301960784313725, 1}$$

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
TeX: {
  Macros: {
    orange: ["{\\color{orange}{#1}}", 1],
    blue: ["{\\color{blue}{#1}}", 1],
    pink: ["{\\color{pink}{#1}}", 1]
  },
  loader: {load: ['[tex]/color']},
  tex: {packages: {'[+]': ['color']}}
}
});
</script>

<style>
.orange {color: #FF9A4D;}
.blue {color: #4D94FF;}
.pink {color: #F94DFF;}
</style>

```{r flair_color, echo=FALSE}
library(flair)
orange <- "#FF9A4D"
blue <- "#4D94FF"
pink <- "#F94DFF"
```

# Extensions of the Linear Model

**Why?**

Standard linear regression model provides interpretable results and works quite well on many real-world problems

However, using such a model makes strong assumptions:

The relationship between predictors and response are

- Additive
- Linear

---

# Removing the additive assumption

Let's consider the linear regression model

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 + \varepsilon$$

This model does only have additive effects

---

# Removing the additive assumption

One way of relaxing the additive assumption and allow for an interaction term is by

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2 + \varepsilon$$

This interaction effect enables the following

$$\begin{align}Y &= \beta_0 + (\beta_1 + \beta_3 X_2) X_1 + \beta_2 X_2 +\varepsilon \\Y &= \beta_0 + \tilde\beta_1 X_1 + \beta_2 X_2 + \varepsilon \end{align}$$

---

# Removing the additive assumption

Now the effect of $X_1$ is no longer constant

Adjusting $X_2$ will change the impact of $X_1$ on $Y$.

It is sometimes the case that an interaction term has a very small p-value but the main associated effects do not

---

# Hierachical Principle

If we include an interaction term in a model, we should also include the main effects, even if the p-values associated with their coefficients are not significant

If the p-value associated with $X_1$ and $X_2$ are not very small we should not worry and we should include them if the p-value associated with $X_1 X_2$ is very small

---

# Non-linear relationships

The reality may be that the relationship between the response and predictors is non-linear

One of the ways we have looked at is to do polynomial regression

---

# Polynomial Regression

Turning

$$Y = \beta_0 + \beta_1 X + \varepsilon$$

into

$$Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \varepsilon$$

Which is extensible to

$$Y = \beta_0 + \beta_1 X + \beta_2 X^2 + ... + \beta_p X^p + \varepsilon$$

(although we rarely use $p > 3$)

---

# Polynomial Regression 

Matrix notation

$$\mathbf Y = \begin{bmatrix} Y_1\\ Y_2\\ \vdots\\ Y_n \end{bmatrix} \qquad \mathbf Y =  \begin{bmatrix}  1   & X_1 & X_1^2 & \dots & X_1^p\\ 1   & X_2 & X_2^2 & \dots & X_2^p\\  \vdots & \vdots & \vdots & & \vdots\\  1   & X_n & X_n^2 & \dots & X_n^p \end{bmatrix} \begin{bmatrix} \beta_1\\ \beta_2\\ \vdots\\ \beta_p \end{bmatrix} + \begin{bmatrix} \varepsilon_1\\ \varepsilon_2\\ \vdots\\ \varepsilon_n \end{bmatrix}$$

---

# Polynomial Regression

Assuming $n>p$, since $\mathbf X$ is a [Vandermonde matrix](https://en.wikipedia.org/wiki/Vandermonde_matrix), the invertibility condition is guaranteed to hold if all the $X_i$ values are distinct and we get a unique least-squares solution

---

# Potential Problems

When we fit a linear regression model to a particular data set, many problems may occur

---

# Non-normality

The Relationship between $Y$ and $X$ is not linear


.pull-left[
### Indicator

- QQ-plot
- Shapiro–Wilk test
- Skewness, Kurtosis
- Histogram; Boxplot

]

.pull-right[
### Remedy

Transformations

or

non-parametric metrics
]

---

# Heteroscedasticity

Non-equal variance

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)

tibble(
  x = runif(100, 5, 50)
) %>%
  mutate(y = runif(100, -(x), (x))/25) %>%
  ggplot(aes(x, y)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 0) +
  theme_minimal() +
  labs(x = "Fitted values", y = "Residuals")
```

---

# Heteroscedasticity

.pull-left[
### Indicator

- Residual plots
- Levene
- Breusch-Pagan test
]

.pull-right[
### Remedy

Apply transformations to $Y$, such as $\log Y$ or $\sqrt Y$

or do weighted least squares
]

---

# Linearity

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)

tibble(
  x = runif(100, -5, 5)
) %>%
  mutate(y = x ^ 2 + rnorm(n(), sd = 5)) %>%
  ggplot(aes(x, y)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 0) +
  theme_minimal() +
  labs(x = "Fitted values", y = "Residuals")
```

---

# Linearity

.pull-left[
### Indicator

- Residual plots
- Lack-of-fit test
]

.pull-right[
### Remedy

- Add predictors
- Use non-linear transformation of the predictors such as $\log Y$, $\sqrt Y$, or $X^2$

]

---

# Indepence

Correlation of error terms

The error terms $\varepsilon_1, \varepsilon_2, ..., \varepsilon_n$ should be correlated

if they are correlated then we may have an unwarranted sense of confidence in our model (narrower confidence bands)

---

# Indepence

.pull-left[
### Indicator

- Residual plots, look for trends
]

.pull-right[
### Remedy

- Fit time series models
- Improve experimental design

]

---

# Outliers / high leverage point

An **outlier** is a point for which $Y_i$ is far from the value predicted by the model

A **high leverage point** point is a point that has extreme predictor values

high leverage observations tend to have a sizeable impact on the estimated regression line

---

# Outliers / high leverage point

.pull-left[
### Indicator

- Residual plots
- Studentized residuals plots

]

.pull-right[
### Remedy

- Find the reason why they are the way they are
- Delete or reweight (you need a good reason to do this)

]

---

# Collinearity

Two or more predictor variables are closely related to one another

---

# Collinearity

### Indicator

Look at the correlation matrix 

Access multicollinearity by computing the variance inflation factor (VIF)

VIF is the ratio of the variance of $\hat \beta_j$ when fitting full model divided by variance of $\hat \beta_j$ if fit on its own

The smallest value is 1 which is great

if VIF > 10 we have problems

---

# Collinearity

### Remedy

- Variable selection
- Ridge regression


