---
title: "Quantitatiive & Qualitative Predictors"
subtitle: "AU STAT-615"
author: "Emil Hvitfeldt"
date: "2021-03-24"
output:
  xaringan::moon_reader:
    css: ["theme.css", "default"]
    lib_dir: libs
    nature:
      beforeInit: "macros.js"
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: [center, middle]
---

```{r include=FALSE}
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
lines <- options$output.lines
if (is.null(lines)) {
  return(hook_output(x, options))  # pass to default hook
}
x <- unlist(strsplit(x, "\n"))
more <- "..."
if (length(lines) == 1) {        # first n lines
  if (length(x) > lines) {
    # truncate the output, but add ....
    x <- c(head(x, lines), more)
  }
} else {
  x <- c(more, x[lines], more)
}
# paste these lines together
x <- paste(c(x, ""), collapse = "\n")
hook_output(x, options)
})
knit_hooks$set(output = function(x, options) {
# this hook is used only when the linewidth option is not NULL
if (!is.null(n <- options$linewidth)) {
  x = knitr:::split_lines(x)
  # any lines wider than n should be wrapped
  if (any(nchar(x) > n)) x = strwrap(x, width = n)
  x = paste(x, collapse = '\n')
}
hook_output(x, options)
})

opts_chunk$set(
echo = TRUE,
fig.width = 7, 
fig.align = 'center',
fig.asp = 0.618, # 1 / phi
out.width = "700px"
)
```

```{r, echo = FALSE, message=FALSE}
library(sass)
sass(sass_file("theme.sass"), output = "theme.css")
library(tibble)
library(dplyr)
library(ggplot2)
```

$$\require{color}\definecolor{orange}{rgb}{1, 0.603921568627451, 0.301960784313725}$$
$$\require{color}\definecolor{blue}{rgb}{0.301960784313725, 0.580392156862745, 1}$$
$$\require{color}\definecolor{pink}{rgb}{0.976470588235294, 0.301960784313725, 1}$$

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
TeX: {
  Macros: {
    orange: ["{\\color{orange}{#1}}", 1],
    blue: ["{\\color{blue}{#1}}", 1],
    pink: ["{\\color{pink}{#1}}", 1]
  },
  loader: {load: ['[tex]/color']},
  tex: {packages: {'[+]': ['color']}}
}
});
</script>

<style>
.orange {color: #FF9A4D;}
.blue {color: #4D94FF;}
.pink {color: #F94DFF;}
</style>

```{r flair_color, echo=FALSE}
library(flair)
orange <- "#FF9A4D"
blue <- "#4D94FF"
pink <- "#F94DFF"
```

# Polynomial Regression Models

One predictor variable - second order

$$Y_i = \beta_0 + \beta_1 X_i + \beta_2 X_i^2 + \varepsilon_i$$

where $X_i = X_i - \bar X$

We are centering the predictor variable because $X$ and $X^2$ may be highly correlated and thus $\mathbf{X}^T\mathbf{X}$ will be very difficult to invert. This can lead to computational issues

---

# Notation

Most of the time we use the following notation

$$Y_i = \beta_0 + \beta_1 X_i + \beta_{11} X_i^2 + \varepsilon_i$$

and

$$E\{Y\} = \beta_0 + \beta_1 X_i + \beta_{11} X_i^2$$

This is done to put emphasize on the exponents

- $\beta_1 \rightarrow$ linear effect coefficient 
- $\beta_{11} \rightarrow$ quadratic effect coefficient

---

# $E\{Y\} = 52 + 8x-2x^2$

```{r, echo=FALSE}
ggplot() +
  xlim(0, 10) +
  geom_function(fun = ~ 52 + 8 * .x - 2 * .x^2) +
  theme_minimal()
```

---

# $E\{Y\} = 18 -8x+2x^2$

```{r, echo=FALSE}
ggplot() +
  xlim(0, 10) +
  geom_function(fun = ~ 18 - 8 * .x + 2 * .x^2) +
  theme_minimal()
```

---

#  One predictor - third orders

$$Y_i = \beta_0 + \beta_1 X_i + \beta_{11} X_i^2 + \beta_{111} X_i^3 + \varepsilon_i$$

where $X_i = X_i - \bar X$

---

# $E\{Y\} = 2 + 7x+5x^2 + x^3$

```{r, echo=FALSE}
ggplot() +
  xlim(-10, 10) +
  geom_function(fun = ~ 2 + 7 * .x +5 * .x^2 + .x^3) +
  theme_minimal()
```

---

# Polynomial regression model

Models can become more complicated. For instance we can consider 

two predictor variables - second order

$$Y_i = \beta_0 + \orange{\beta_1 X_{i1}} + \blue{\beta_2 X_{i2}} + \orange{\beta_{11} X_{i1}^2} + \blue{\beta_{22} X_{i2}^2} + \beta_{12} X_{i1} X_{i2}$$
---

# Interaction Regression models

Let us have $p-1$ predictor variables. A regression model containns .blue[additive effects] if the response function can be written in the form

$$E\{ Y \} f_1(X_1) + f_2(X_2) + \cdots + f_{p-1} (X_{p-1})$$

---

# Interaction Regression models

Example

$$E\{Y\} = \blue{\beta_0 + \beta_1 X_1 + \beta_2 X_1^2} + \pink{\beta_3 X_2}$$

- $\blue{f_1(X_1)}$
- $\pink{f_2(X_2)}$

---

# Interaction Regression models

On the other hand, if we have

$$E\{Y\} = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2$$

Cannot be expressed in the previous form

This model contains .blue[interactionn effects]

---

# Interaction Regression models

On the other hand, if we have

$$E\{Y\} = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \orange{\beta_3 X_1 X_2}$$

.orange[This] is called linear-by-linear or bilinear interaction ter or simply interraction term

---

# Interpretation

The regressionn model for two quanntitative predictor variables with linear effects on $Y$ and interacting effect on $X_1$ and $X_2$ on $Y$ represented by a cross product is as follows

$$Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \beta_3 X_{i1} X_{i2} + \varepsilon_i$$

---

# Interpretation

Note: The regressionn coefficients $\beta_1$ and $\beta_2$ no longer indicate the change in the mean response with a unit increase of the predictor variable  with the other predictor variable held constant at any given level

It can be shown that the change in the mean response with a unit increase in $X_1$ when $X_2$ is held constant is 

$$\beta_1 + \beta_2 X_2$$

---

# Qualitative Predictors

Example of qualitative predictors

$$X_2 = \begin{cases} 1 & \text{If stock company}\\ 0 & \text{Otherwise} \end{cases}$$

$$X_3 = \begin{cases} 1 & \text{If mutual company}\\ 0 & \text{Otherwise} \end{cases}$$

---

# Qualitative Predictors

In order to define the qualitative variables, we used indicator functions and generate the .blue[indicator variables] or .blue[dummy variable]

---

# Qualitative Predictors

Let

$$Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \beta_3 X_{i3} + \varepsilon_i$$

Where $Y$ indicates the speed with which a particular insurance innovation is adopted

$X_1$ is the size of the firm annd $X_2$ and $X_3$ indicate the type of firm

---

# Qualitative Predictors

Let us assume that we have $n=4$ observations

$$\mathbf{X} = \begin{bmatrix} 1 & X_{11} & 1 & 0 \\ 1 & X_{21} & 1 & 0 \\ 1 & X_{31} & 0 & 1 \\ 1 & X_{41} & 0 & 1  \end{bmatrix}$$

---

# Qualitative Predictors

Note that 

$$\begin{bmatrix} 1\\ 1\\ 1\\ 1 \end{bmatrix} = 1 \cdot \begin{bmatrix} 1\\ 1\\ 0\\ 0 \end{bmatrix}  + 1 \cdot \begin{bmatrix} 0\\ 0\\ 1\\ 1 \end{bmatrix}$$

Which implies that the columns are linearly dependent

Thus we cannot invert $\mathbf{X}^T \mathbf{X}$, so we cannot have unique solutions for the estimator.

.pink[Solution:] Drop one of the indicator variables

---

# Qualitative Predictors

Note: A qualitative variable with $c$ classes will be represented by $c-1$ indicator variables, each taking on the values $0$ and $1$.

---

# Interpretationn of Regression coefficents

Suppose that we drop the indicator variable $X_3$ from the model

Then we have

$$Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \varepsilon_i$$

**Case 1:** Mutual firms

$$E\{Y\} = \beta_0 + \beta_1 X_1$$

**Case 2:** Stock firms

$$E\{Y\} = \beta_0 + \beta_1 X_{1} + \beta_2 \cdot 1$$

---

# Interpretationn of Regression coefficents

The critical question here is why we do not simply fit seperate regressions for stock firms and mutual forms and instead we adopted the approach of fitting one regression with an indicator variable

### Reason 1

Since the model assumes equal slopes and same constant error term variancee for each type of firm, the common slope $\beta_1$ can best bbe estimated by pooling the two types of firms

---

# Interpretationn of Regression coefficents

The critical question here is why we do not simply fit seperate regressions for stock firms and mutual forms and instead we adopted the approach of fitting one regression with an indicator variable

### Reason 2

Other inferences such as for $\beta_0$ and $\beta_2$ can be made more precisely by working with one regression model containing an indicator variable since more degrees of freedom will be associated with MSE

We want a small MSE, so we nneed to devide by more degrees of freedom

---

# Qualitative Predictors

If a qualitative variable has more than two classes, we requiree additionnal indicator variables in the regressionn model

$$X_2 = \begin{cases} 1 & \text{If } M_1\\ 0 & \text{Otherwise} \end{cases}$$

$$X_3 = \begin{cases} 1 & \text{If } M_2\\ 0 & \text{Otherwise} \end{cases}$$

$$X_4 = \begin{cases} 1 & \text{If } M_3\\ 0 & \text{Otherwise} \end{cases}$$

And we can work in the same way we did previously

---

# Alternatives to indicator variables

Consider the following table

| Class         | $X_1$ |
|---------------|-------|
| Frequent user | 3     |
| Occasional    | 2     |
| Non user      | 1     |

---

# Alternatives to indicator variables

The allocated codes that define the metric may not be reasonable as a quantitative variable

The mean response would change by the same amount when going from a non user to an occasional user as when going from a occasional user to a frequent user

---

# Indicator variables

Indicator variables can be used even if the predictor variable is quantitative

For example

If we have data regarding ages of people, then we can arrrange the groups such as

- under 21
- 21-34
- 35-49
- 50-65
- over 65

---

# Indicator variables


$$X_2 = \begin{cases} 1 & \text{If stock company}\\ -1 & \text{If mutual company} \end{cases}$$

here a meaningful test will be $H_0: \beta_2 = 0$ vs $H_\alpha: \beta_2 \neq 0$

since the two sides would be equal to each other when $\beta_2 = 0$

---

### Inteeractionn between qualitative and quantitative predictors

For example

$$\begin{align} X_{i1} &= \text{size of firm}\\ X_{i2} &= \begin{cases} 1 & \text{If stock company}\\ 0 & \text{otherwise} \end{cases}\end{align}$$

We can have

$$
\begin{align}
Y_i &= \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \beta_3 X_{i1} X_{i2} + \varepsilon_i \\
E\{Y\} &= \beta_0 + \beta_1 X_{1} + \beta_2 X_{2} + \beta_3 X_{1} X_{2}\\
\end{align}
$$
---

### Inteeractionn between qualitative and quantitative predictors

For mutual firm ( $X_2 = 0$ )

$$E\{Y\} = \beta_0 + \beta_1 X_{1}$$

For stock firm ( $X_2 = 0$ )

$$E\{Y\} = \beta_0 + \beta_1 X_{1} + \beta_2 + \beta_3 X_{1} = (\beta_0 + \beta_2) + (\beta_1 + \beta_3) X_1$$

